---
title: "Reproducible Research: Peer Assessment 1"
author: "Arnulfo Díaz Ulate"
output: 
  html_document:
    keep_md: true
---


```{r initial setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

<br><br>

### 1. Loading and preprocessing the data:
```{r}
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", "activity_data.zip")
unzip("activity_data.zip")
dt<- read.csv("activity.csv", stringsAsFactors = F)
```

Install ggplot2, and sqldf library 
```{r ggplot2, echo = FALSE}
# install lubridate for date analysis
x = 0
tryCatch( expr = {
        library(ggplot2)
        x = 1
    }
)
if (x == 0) {
    install.packages("ggplot2")
    library(ggplot2)
}
```
```{r sqldf, echo = FALSE}
x = 0
tryCatch( expr = {
        library(sqldf)
        x = 1
    }
)
if (x == 0) {
    install.packages("sqldf")
    library(sqldf)
}
```

Remove the missing values
```{r}
missing_dt <- dt[is.na(dt$steps),]
dim(missing_dt)
dt1 <- dt[!is.na(dt$steps),]
```

<br><br>

### 2. What is mean total number of steps taken per day?

#### Histogram:
```{r}
StepsPerDay <- sqldf("Select date,sum(Steps) as TotalSteps from dt1 group by date")
hist(StepsPerDay$TotalSteps, main = "Histogram of total number of steps taken per day", xlab = "Total number of steps")
```

#### mean and median:

**Info** of steps taken each day:
```{r}
summary(StepsPerDay)
```
<br><br>

### 3. What is the average daily activity pattern?

#### Time series plot:
```{r}
AvgStepsbyInterval <- sqldf("Select interval,avg(Steps) as AvgSteps from dt1 group by interval")
plot(AvgStepsbyInterval$interval,AvgStepsbyInterval$AvgSteps, type = "l", main = "Time series plot of the \n average number of steps taken", xlab = "interval", ylab = "Mean steps")
```

#### The 5-minute interval contains the max number of steps:
```{r}
MaxInterval<- sqldf("Select interval,max(AvgSteps) from AvgStepsbyInterval")
MaxInterval
```
<br><br>

### 4. Imputing missing values:

In this case **mean of the intervals** is used
```{r}
MeanSteps<-sqldf("Select avg(AvgSteps) from AvgStepsbyInterval")
missing_dt$steps <-MeanSteps$`avg(AvgSteps)`
```

assign avg to all NA in a new dataset:
```{r}
new_dt <- sqldf("Select * from dt1
                union all
                Select * from missing_dt")
```

#### Histogram after missing values are imputed:
```{r}
New_StepsPerDay <- sqldf("Select date,sum(Steps) as TotalSteps from new_dt group by date")

hist(New_StepsPerDay$TotalSteps, main = "Histogram of total number of steps taken per day", xlab = "Total number of steps")
```

<br><br>

### 5. Differences in activity patterns between weekdays and weekends:

Panel plot:
```{r}
summary(New_StepsPerDay)
summary(StepsPerDay)
```

#### New Factor of weekdays and weekends
```{r}
new_dt$days <- weekdays(as.Date(new_dt$date))
weekend <- c("Sunday","Saturday")
weekday <- c("Monday","Tuesday","Wednesday","Thursday","Friday")
new_dt$days[new_dt$days %in% weekend] <- "Weekend"
new_dt$days[new_dt$days %in% weekday] <- "weekday"
```

#### Difference between weekdays and weekends
```{r}
mean_number_steps <- aggregate(steps~ interval+days, new_dt, mean)
g <- qplot(interval, steps, data = mean_number_steps, facets = days~.)
g + geom_line(size = 1) + ylab("Mean steps") + ggtitle("Average number of steps taken, \n averaged across all weekday days or weekend days ")
```